---
title: "Grand Rapids Crashes"
author: "MADAN"
date: "4/9/2022"
output: html_document
---

```{r}
setwd("~/STA 518/STA Final Project/STA518-Final-Project")
```

```{r}
library(readr)
total_crashes <- read_csv("total_crashes.csv")
total_crashes
```


```{r}
library(readr)
alcohol_crashes <- read_csv("alcohol_crashes.csv")
alcohol_crashes
```

```{r}
library(readr)
drug_crashes <- read_csv("drug_crashes.csv")
drug_crashes
```


```{r}
library(readr)
alcohol_injury_death <- read_csv("alcohol_injury_death.csv")
alcohol_injury_death
```


```{r}
library(readr)
drug_deaths <- read_csv("drug_injury_death.csv")
drug_deaths
```

Merging datasets by year
```{r}
df = merge(x = total_crashes, y = alcohol_crashes, by = "year")
df = merge(x = df, y =alcohol_injury_death , by = "year")
df = merge(x = df, y = drug_crashes, by = "year")
df = merge(x = df, y = drug_deaths, by = "year")
#dropping the row of 2020 as it has multiple NA
df = df[-17,]
df
```

Adding columns
Total crashes:
```{r}
df$total_crash_sum <- rowSums(df[,c("Ionia_total", "Kent_total","Montcalm_total", "Ottawa_total")])
df
```

Total alcohol crashes
```{r}
df$total_alcohol_crashes<- rowSums(df[,c("Ionia_alcohol_crash", "Kent_alcohol_crash","Montcalm_alcohol_crash", "Ottawa_alcohol_crash")])
df
```

Total of alcohol injuiries and death
```{r}
df$total_alcohol_injuries  <- rowSums(df[,c("Ionia_alcohol_injury", "Kent_alcohol_injury","Montcalm_alcohol_injury", "Ottawa_alcohol_injury")])
df
```

Drug crashes
```{r}
df$total_drug_crashes  <- rowSums(df[,c("Ionia_dcrashes", "Kent_dcrashes","Montcalm_dcrashes", "Ottawa_dcrashes")])
df
```

Drug injuries and alcohol
```{r}
df$total_drug_deaths  <- rowSums(df[,c("Ionia_injury_deaths", "Kent_injury_deaths","Montcalm_injury_deaths", "Ottawa_injury_deaths")])
df
```

Visualizations:
1. Bar chart of total injuries and death due to drugs and alcohol library(reshape2)
```{r}
library(reshape2)
dfm <- melt(df[,c('year','total_alcohol_injuries','total_drug_deaths')], id.vars = 1)
ggplot(dfm,aes(x = year, y = value)) + 
  geom_bar(aes(fill = variable),stat = "identity",position = "dodge")
```

2. Line chart for crashes between counties

2.1 Line chart of injuries and death due to alcohol

```{r}
county_alcohol_crashes <- data.frame(year = df$year,                            # Reshape data frame
                       alcrashes = c(df$Montcalm_alcohol_crash, df$Ottawa_alcohol_crash, df$Ionia_alcohol_crash,df$Kent_alcohol_crash),
                       group = c(rep("Montcalm_alcohol_crash", nrow(df)),
                                 rep("Ottawa_alcohol_crash", nrow(df)),
                                 rep("Ionia_alcohol_crash", nrow(df)),
                                 rep("Kent_alcohol_crash", nrow(df))))


library("ggplot2")
ggp <- ggplot(county_alcohol_crashes, aes(year, alcrashes, col = group)) +             # Create ggplot2 plot
  geom_line()
ggp 
```
2.2 Injuries and deaths in four counties due to alcohol

```{r}
county_alcohol_injury_death <- data.frame(year = df$year,                            # Reshape data frame
                       alcohol_injury_and_death = c(df$Montcalm_alcohol_injury, df$Ottawa_alcohol_injury, df$Ionia_alcohol_injury,df$Kent_alcohol_injury),
                       group = c(rep("Montcalm_alcohol_injury", nrow(df)),
                                 rep("Ottawa_alcohol_injury", nrow(df)),
                                 rep("Ionia_alcohol_injury", nrow(df)),
                                 rep("Kent_alcohol_injury", nrow(df))))


library("ggplot2")
ggp <- ggplot(county_alcohol_injury_death, aes(year, alcohol_injury_and_death, col = group)) +             # Create ggplot2 plot
  geom_line()
ggp
```

2.3 Crashes in four Counties due to drug
```{r}
county_drug_crashes <- data.frame(year = df$year,                            # Reshape data frame
                       drug_crashes = c(df$Montcalm_dcrashes, df$Ottawa_dcrashes, df$Ionia_dcrashes,df$Kent_dcrashes),
                       group = c(rep("Montcalm_dcrashes", nrow(df)),
                                 rep("Ottawa_dcrashes", nrow(df)),
                                 rep("Ionia_dcrashes", nrow(df)),
                                 rep("Kent_dcrashes", nrow(df))))


library("ggplot2")
ggp <- ggplot(county_drug_crashes, aes(year, drug_crashes, col = group)) +             # Create ggplot2 plot
  geom_line()
ggp 
```

2.4 Injuries and deaths in four counties due to drugs
```{r}
county_drug_injury_death <- data.frame(year = df$year,                            # Reshape data frame
                       alcohol_injuries_deaths = c(df$Montcalm_injury_deaths, df$Ottawa_injury_deaths, df$Ionia_injury_deaths,df$Kent_injury_deaths),
                       group = c(rep("Montcalm_injury_deaths", nrow(df)),
                                 rep("Ottawa_injury_deaths", nrow(df)),
                                 rep("Ionia_injury_deaths", nrow(df)),
                                 rep("Kent_injury_deaths", nrow(df))))


library("ggplot2")
ggp <- ggplot(county_drug_injury_death, aes(year, alcohol_injuries_deaths, col = group)) +             # Create ggplot2 plot
  geom_line()
ggp
```


```{r}
```

3. Pie chart
3.1 Proportion of crashes due to Drugs, alcohol and other factors

```{r}
df$Other_Crashes <- df$total_crash_sum - df$total_alcohol_crashes - df$total_drug_crashes #Crashes due to factors other than alcohol and drugs
df <- data.frame(
  Types = c("alcohol", "drug", "others"),
  value=c("total_alcohol_crashes","total_drug_crashes","Other_Crashes")
)
ggplot(df, aes(x="", y=value, fill=Types)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  theme_void()
```




